---
title: "TMS-Assayed Saliva Stability Analysis"
author: "Alex Spiers"
date: "08/11/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(brms)
library(knitr)
library(kableExtra)
library(lme4)
library(lmerTest)
library(jtools)
```

## Explanation of modelling process

This R Markdown document estimates the degradation of steroids in long-term
storage. Samples were extracted from research subjects and analysed at 
six different timepoints, with the last timepoint >600 days after baseline.

```{r process_data}

source("./process_tms_data.R")
# REMOVE OUTLIER
tms_processed <- tms_processed %>% filter(fraction < 5)

tms_processed$conc_nmol <- tms_processed$value / 1000
```

### Initial plots of raw data

Plots of raw salivary steroid concentration against time

```{r plot_raw, echo=TRUE}
ggplot(data = tms_processed,
       aes(x=days, y=value, col=sample_id, group=sample_id)) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = "none") +
  facet_wrap(vars(biomarker), scales = "free")
```

### Initial plots of transformed (scaled as proportion of baseline) data

A simple spline model is fitted to estimate general trend

```{r plot_proportions, echo=TRUE}
ggplot(data = tms_processed,
       aes(x = days, y = 100 * (fraction - 1))) +
  ylab("% Change") +
  geom_line(aes(col=sample_id, group=sample_id)) +
  theme_minimal() +
  theme(legend.position = "none") +
  facet_wrap(vars(biomarker), scales = "free") +
  stat_smooth(method = "gam", formula = y ~ s(x, k = 4), size = 1)
```

## Mixed effects ANOVA
First we run a mixed effects linear model to test the hypothesis $\beta_1 = 0$ for formula below:

$$conc = beta_0 + u_j + beta_1time + \epsilon$$
where $$u_j \sim N(0, \sigma_j)$$

for subjects $j = 1, ... ,n$

```{r mixed_ANOVA, results='asis', echo=FALSE}
for (bio in unique(tms_processed$biomarker)){
    cat("\n")
    cat(paste("#### Analysing:", bio, "\n"))
    cat("\n")

    df <- tms_processed %>% filter(biomarker == bio) %>%
        mutate(time = years)
    mixed_linear <- lme4::lmer(value ~ time + (1 | sample_id), data = df)
    cat("\n")
    cat(paste0("Summary for mixed effect model for", bio, ": ", "\n"))
    cat(kable(as.data.frame(summary(mixed_linear)$coef)) %>% kable_styling())
    cat("\n")
    cat(paste("Confidence intevals for annual degradation as percentage: \n"))
    cat("\n")
    ci_model <- confint(mixed_linear)[4,] %>%
        kbl() %>%
        kable_styling()
    cat(paste(ci_model, "\n"))
}
```

## Estimation of degradation rate

In this section we run 4 models with increasing complexity in order to analyse stability and estimate rate of degradation (and half-life DT50).

Prior to analysis, the data was transformed so that all salivary concentrations
are as a proportion of baseline concentration. i.e.:

$$
y_{transformed} = y_t / y_0
$$

for all $t \sim 1, ... , 5$

For each biomarker we model degradation using four types of likelihood functions:

1. Simple linear model (% as function of time)

$$y = y_0 - kt$$

2. Hierarchical linear model (each participant has different slope)

$$y = y_0 - k_jt$$
$$k \sim N(0, \sigma_j)$$

3. Single-First Order (SFO) Exponential Decay

$$
\frac{dy}{dt} = -ky
$$

i.e. the rate of degradation is proportional to the concentration of biomarker. This translates to the model:

$$
y = y_0e^{-kt}
$$

4. Hierarchical SFO (each participant has different rate parameter)

$$y = y_0e^{-kt}$$
$$ k \sim N(0, \sigma_j)$$

Following this, we fit four models using a Bayesian framework. First we capture available structural knowledge about parameters by setting weakly informative
prior distributions the parameters (except for the intercept, which has a prior centred around $y_0 = 1$).

We then obtain the prior distribution for the parameters

We then use [leave-one-out cross-validation](https://mc-stan.org/loo/reference/psis.html) in order to 

<!-- Only run code chunk if models are NOT all fitted --> 

```{r fitting_models, echo=FALSE, debug=TRUE}

REFIT_SAVED_MODELS = FALSE

biomarkers <- unique(tms_processed$biomarker)

SFO_k_fixed_formula <- brmsformula(
  fraction ~ C0 * exp(-k * years),
  C0~1+(1|sample_id), 
  k~1,
  nl=TRUE)

SFO_k_random_formula <- brmsformula(
  fraction ~ C0 * exp(-k * years),
  C0~1+(1|sample_id),
  k~1+(1|sample_id),
  nl=TRUE)

for (bio in biomarkers){
    print("Calculating stability for")
    print(bio)

    df <- tms_processed %>% filter(biomarker == bio)

    #SET PRIORS FOR linear models
    location_b <- mean(df$fraction) %>% signif(., 2)
    scale_b <- sd(df$fraction)

    b_class_prior <- paste0("normal(0, 1)")
    intercept_prior <- paste0("normal(1, 0.02)")
    sigma_prior <- paste0("student_t(4, 0, 0.2)")
    group_variance_prior <- paste0("student_t(4, 0, 0.1)")

    prior_linear_mod1 <- c(
        prior_string(b_class_prior, class="b"),
        prior_string(intercept_prior, class="Intercept"),
        prior_string(group_variance_prior, class="sd"),
        prior_string(sigma_prior, class="sigma")
        )
    if (!file.exists(paste0("./models/", bio, "_linear_mod1.rds")) | REFIT_SAVED_MODELS){
        print("Fitting linear model (fixed k)")
        linear_mod1 <- brm(
            formula = fraction ~ years + (1|sample_id),
            data = df,
            prior = prior_linear_mod1,
            #sample_prior = TRUE,
            family = gaussian(),
            chains = 6,
            cores = 12,
            iter = 3000,
            warmup = 1000,
            #backend = "cmdstanr",
            save_pars = save_pars(all = TRUE),
            control=list(adapt_delta=0.99, max_treedepth=15)
            )
        linear_mod1 <- add_criterion(linear_mod1, "loo", moment_match = TRUE)
        saveRDS(
            linear_mod1, 
            paste0("./models/", bio, "_linear_mod1.rds")
            )
        Sys.sleep(5)
    } else {
        print("Loading from file linear model (fixed k)")
        linear_mod1 <- readRDS(paste0("./models/", bio, "_linear_mod1.rds"))
    }
    
    print(summary(linear_mod1))

    prior_linear_mod2 <- c(
        prior_string(b_class_prior, class = "b"),
        prior_string(intercept_prior, class = "Intercept"),
        prior_string(group_variance_prior, class = "sd"),
        prior_string(sigma_prior, class = "sigma"),
        prior(lkj(2), class = "cor")
    )

    print("Fitting linear model (k varies between subject)")

    if (!file.exists(paste0("./models/", bio, "_linear_mod2.rds")) | REFIT_SAVED_MODELS){
    linear_mod2 <- brm(
        formula = fraction ~ years + (years | sample_id),
        data = df,
        prior = prior_linear_mod2,
        #sample_prior = TRUE,
        family = gaussian(),
        chains = 6,
        cores = 12,
        iter = 3000,
        warmup = 1000,
        #backend = "cmdstanr",
        save_pars = save_pars(all = TRUE),
        control=list(adapt_delta=0.99, max_treedepth=15)
    )
        linear_mod2 <- add_criterion(linear_mod2, "loo", moment_match=TRUE)
        saveRDS(
            linear_mod2, 
            paste0("./models/", bio, "_linear_mod2.rds"))
    Sys.sleep(5)
    } else {
        print("Loading from file linear model (between-subject k varies)")
        linear_mod2 <- readRDS(paste0("./models/", bio, "_linear_mod2.rds"))   
    }
  
    print(summary(linear_mod2))
    #Set priors for SFO models
    prior_SFO1 <- c(
        prior_string(intercept_prior, class="b", nlpar="C0"),
        prior_string("normal(0, 1)", lb=0, nlpar="k"),
        prior_string(group_variance_prior, class="sd", nlpar="C0"),
        prior_string(sigma_prior, class="sigma")
        )

    if (!file.exists(paste0("./models/", bio, "_SFO_model1.rds")) | REFIT_SAVED_MODELS){
        print("Fitting SFO model (fixed k)")
        SFO_model1 <- brm(
            formula = SFO_k_fixed_formula,
            #sample_prior = TRUE,
            data = df,
            family=gaussian(),
            prior = prior_SFO1,
            chains = 6,
            cores = 12,
            iter = 3000,
            warmup = 1000,
            #backend = "cmdstanr",
            save_pars = save_pars(all = TRUE),
            control=list(adapt_delta=0.99, max_treedepth=15)
            )
        SFO_model1 <- add_criterion(SFO_model1, "loo", moment_match=TRUE)
        saveRDS(
            SFO_model1,
            paste0("./models/", bio, "_SFO_model1.rds"))

        Sys.sleep(5)
    } else {
        print("Loading from file SFO model (fixed k)")
        SFO_model1 <- readRDS(paste0("./models/", bio, "_SFO_model1.rds"))   
    }

    print(summary(SFO_model1))

    prior_SFO2 <- c(
        prior_string(intercept_prior, class="b", nlpar="C0"),
        prior_string("normal(0, 1)", lb=0, nlpar="k"),
        prior_string(group_variance_prior, class="sd", nlpar="k"),
        prior_string(group_variance_prior, class="sd", nlpar="C0"),
        prior_string(sigma_prior, class="sigma")
    )
    if (!file.exists(paste0("./models/", bio, "_SFO_model2.rds")) | REFIT_SAVED_MODELS){
        print("Fitting SFO model (k varies between subject)")
        SFO_model2 <- brm(
            formula = SFO_k_random_formula,
            #sample_prior = TRUE,
            data = df,
            family=gaussian(),
            prior = prior_SFO2,
            chains = 6,
            cores = 12,
            iter = 3000,
            warmup = 1000,
            #backend = "cmdstanr",
            save_pars = save_pars(all = TRUE),
            control=list(adapt_delta=0.99, max_treedepth=15)
        )
        SFO_model2 <- add_criterion(SFO_model2, "loo", moment_match=TRUE)
        saveRDS(
            SFO_model2,
            paste0("./models/", bio, "_SFO_model2.rds")
        )
        Sys.sleep(5)
    } else {
        print("Loading from file SFO model (k varies between subject)")
        SFO_model2 <- readRDS(paste0("./models/", bio, "_SFO_model2.rds"))   
    }

    print(summary(SFO_model2))

    print("comparing the LOO for models for")
    print(bio)
    print(loo_compare(linear_mod1, linear_mod2, SFO_model1, SFO_model2))
    print("completed analysis for")
    print(bio)
}
```

<!-- Only run this section if models are already fitted --> 
