---
title: "TMS-Assayed Saliva Stability Analysis"
author: "Alex Spiers"
date: "08/11/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(brms)
```

## Explanation of modelling process

This R Markdown document estimates the degradation of steroids in long-term
storage. Samples were extracted from reesarch subjects and analysed at 
six different timepoints, with the last timepoint >600 days after baseline.

Prior to estimation, the data was transformed so that all salivary concentrations
are as a proportion of baseline concentration. i.e.:

$$
y_{transformed} = y_t / y_0
$$

for all $t \sim 1, ... , 5$



```{r process_data}

source("./process_tms_data.R")
# REMOVE OUTLIER
tms_processed <- tms_processed %>% filter(fraction < 5)

tms_processed$conc_nmol <- tms_processed$value / 1000
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
