---
title: "Affinity-Assayed Saliva Stability Analysis"
author: "Alex Spiers"
date: "08/11/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(brms)
library(knitr)
library(kableExtra)
library(lme4)
library(lmerTest)
library(jtools)
```

## Explanation of modelling process

This R Markdown document estimates the degradation of steroids in long-term
storage. Samples were extracted from research subjects and analysed at 
six different timepoints, with the last timepoint >600 days after baseline.

```{r process_data}

source("./process_affinity_data.R")
# REMOVE OUTLIER
affinity_steroids <- affinity_steroids %>% filter(fraction < 5)

affinity_steroids$conc_nmol <- affinity_steroids$value / 1000
```

### Initial plots of raw data

Plots of raw salivary steroid concentration against time

```{r plot_raw, echo=TRUE}
ggplot(data = affinity_steroids,
       aes(x=days, y=value, col=sample_id, group=sample_id)) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = "none") +
  facet_wrap(vars(biomarker), scales = "free")
```

### Initial plots of transformed (scaled as proportion of baseline) data

A simple spline model is fitted to estimate general trend

```{r plot_proportions, echo=TRUE}
ggplot(data = affinity_steroids,
       aes(x = days, y = 100 * (fraction - 1))) +
  ylab("% Change") +
  geom_line(aes(col=sample_id, group=sample_id)) +
  theme_minimal() +
  theme(legend.position = "none") +
  facet_wrap(vars(biomarker), scales = "free") +
  stat_smooth(method = "gam", formula = y ~ s(x, k = 4), size = 1)
```

## Mixed effects ANOVA
First we run a mixed effects linear model to test the hypothesis $\beta_1 = 0$ for formula below:

$$conc = beta_0 + u_j + beta_1time + \epsilon$$
where $$u_j \sim N(0, \sigma_j)$$

for subjects $j = 1, ... ,n$

```{r mixed_ANOVA, results='asis', echo=FALSE,eval=TRUE}
for (bio in unique(affinity_steroids$biomarker)){
    cat("\n")
    cat(paste("#### Analysing:", bio, "\n"))
    cat("\n")

    df <- affinity_steroids %>% filter(biomarker == bio) %>%
        mutate(time = years)
    mixed_linear <- lme4::lmer(fraction ~ time + (1 | sample_id), data = df)
    cat("\n")
    cat(paste0("Summary for mixed effect model for", bio, ": ", "\n"))
    cat(kable(as.data.frame(summary(mixed_linear)$coef)) %>% kable_styling())
    cat("\n")
    cat(paste("Confidence intevals for annual degradation as percentage: \n"))
    cat("\n")
    ci_model <- confint(mixed_linear)[4,] %>%
        kbl() %>%
        kable_styling()
    cat(paste(ci_model, "\n"))
}
```

## Estimation of degradation rate

<!-- Only run code chunk if models are NOT all fitted --> 

```{r fitting_models, echo=FALSE, debug=TRUE}

REFIT_SAVED_MODELS = FALSE

biomarkers <- unique(affinity_steroids$biomarker)

b_class_prior <- paste0("normal(0, 0.5)")
intercept_prior <- paste0("normal(1, 0.02)")
sigma_prior <- paste0("student_t(4, 0, 0.2)")
group_variance_prior <- paste0("student_t(4, 0, 0.1)")
corr_matrix_prior <- "lkj_corr_cholesky(2)"

prior_linear_mod <- c(
  prior_string(b_class_prior, class = "b"),
  prior_string(intercept_prior, class = "Intercept"),
  prior_string(group_variance_prior, class = "sd"),
  prior_string(sigma_prior, class = "sigma"),
  prior_string(corr_matrix_prior, class = "L")
)

for (bio in biomarkers) {
    print("Calculating stability for")
    print(bio)

    df <- affinity_steroids %>% filter(biomarker == bio)


    if (!file.exists(paste0("./models/", bio, "_linear_increase.rds")) | REFIT_SAVED_MODELS){
    linear_mod <- brm(
        formula = fraction ~ years + (years | sample_id),
        data = df,
        prior = prior_linear_mod,
        #sample_prior = TRUE,
        family = gaussian(),
        chains = 6,
        cores = 12,
        iter = 3000,
        warmup = 1000,
        backend = "cmdstanr"#,
        #control=list(adapt_delta=0.99, max_treedepth=15)
    )
    saveRDS(
        linear_mod,
        paste0("./models/", bio, "_linear_increase.rds"))
    Sys.sleep(5)
    }
}

summary(linear_mod)
```

<!-- Only run this section if models are already fitted --> 
